Destruction of metrizability in generalized inverse limits

### 2015-08-21 articles/destroying-metrizability-in-generalized-inverse-limits/article.tex
corrects per Scott's comments

\begin{abstract}
If $X$ is a compact Hausdorff space, an upper semi-continuous bonding function $f: X \rightarrow 2^{X}$ is said to be idempotent if $f^2 = f$. In this short paper, we prove that if $f: [0,1] \rightarrow C([0,1])$ is u.s.c., idempotent and surjective, but $f$ is not the identity, then the inverse limit with the single bonding function $f$ and with factor spaces indexed by the ordinal $\kappa$ contains a copy of $\kappa + 1$. It follows that such an inverse limit is only metric in the case where the index set $\kappa$ is countable.
\end{abstract}

@@

This short paper focuses on the case of an inverse limit with a single idempotent, surjective, upper semi-continuous, continuum-valued bonding function on $[0,1]$, when the index set of the inverse limit is some ordinal $\kappa$. In particular, we show that (so long as the bonding function is not the identity) such an inverse limit is metric exactly when $\kappa$ is countable. The strategy, as will be seen, is to show that the inverse limit must contain a copy of $\kappa + 1$.

@@

Let $\kappa$ be an ordinal. (We follow the convention of set theory in which an ordinal is equal to the set of its predecessors. For more detailed background on ordinals, see, for example, \cite{Kunen}.) Suppose that for each $\alpha \in \kappa$, $X_{\alpha}$ is a non-empty compact Hausdorff space and whenever $\alpha \le \beta \in \kappa$, the set-valued function $f_{\alpha \beta} : X_{\beta} \rightarrow 2^{X_{\alpha}}$ is u.s.c. (With this notation, $f_{\alpha \alpha}$ always denotes the identity on $X_{\alpha}$.) Suppose further that, whenever $\alpha \le \beta \le \eta \in \kappa$, we have $f_{\alpha \beta} \circ f_{\beta \eta} = f_{\alpha \eta}$. Then the collection $\textbf{f} = \{ f_{\alpha \beta} \ | \ \alpha \le \beta \in \kappa\}$ of u.s.c. functions is said to be \emph{exact}. In this case, we say $\{X_{\alpha}, f_{\alpha \beta}, \kappa\}$ is an \emph{inverse limit system}; the \emph{inverse limit}, $\varprojlim \{X_{\alpha}, f_{\alpha \beta}, \kappa\}$, of this system is $\{ \textbf{x} \in \prod_{\alpha \in \kappa} X_{\alpha} \ | \ x_{\alpha} \in f_{\alpha \beta}(x_{\beta})$ for all $\alpha \le \beta \in \kappa\}$. (In keeping with tradition, a boldface $\textbf{x}$ always denotes an element of $\prod_{\alpha \in \kappa} X_{\alpha}$; the $\alpha$th coordinate of such an $\textbf{x}$ will be denoted $\textbf{x}(\alpha)$.)  We call the u.s.c. functions $f_{\alpha \beta}$ the \emph{bonding functions} of the inverse limit, and each $X_{\alpha}$ is called a \emph{factor space} of the inverse limit.

@@

\textbf{Case 3:}
$x\in(0,1)$. Since $f(0)\subseteq(0,1]$ and the graph of $f$ is connected, $0\not\in f([0,x))$. Likewise, $f(1)\subseteq[0,1)$ and $1\not\in f((x,1])$. Since $f$ is surjective, $1\in f(v)$ for some $v\in[0,x]$ and $0\in f(w)$ for some $w\in[x,1]$. Let us say $f(v) = [k,1]$.

@@

So, $f(1) = [a,b]$ where $[a,b] \subseteq [0,x)$ or $[a,b] \subseteq (x,1)$. If $[a,b] \subseteq (x,1)$, then since $a > x$, $f(a) \subseteq [0,a)$ while $f(a)\subseteq f([a,b])=f(f(1))=f(1)=[a,b]$, contradiction. On the other hand, if $[a,b] \subseteq [0,x)$, then since $b<x$, $f(b) \subseteq (b,1]$ while $f(b) \subseteq f([a,b])=f(f(1)=f(1)=[a,b]$, another contradiction. Since each possible case results in a contradiction, the proof is complete.
\end{proof}

@@

\begin{proof}
As $G(f)$ satisfies $\Gamma$, let us say without loss of generality that $\<x,x\>, \<y,y\>$ and $\<x,y\>$ are elements of $G(f)$. $\varprojlim\{[0,1],f,\kappa\}$ contains the points $\textbf{x}_\alpha$ for $\alpha\leq\kappa$, defined by $\textbf{x}_\alpha(\beta)=y$ for $\beta<\alpha$ and $\textbf{x}_\alpha(\beta)=x$ otherwise. It's easy to see that $\{\textbf{x}_\alpha:\alpha\leq\kappa\}$ is a copy of $\kappa+1$.
\end{proof}

@@

Recall that a \textit{Corson compact} space is a compact space which embeds in a $\Sigma$-product of real lines (see \cite{alster}). All compact metric spaces are embeddable in $\mathbb R^\omega$ and therefore Corson compact.

@@

\bibitem{alster} Alster, K., and Pol, Roman. ``On function spaces of compact subspaces of $\Sigma$-products of the real line.'' Fundamenta Mathematicae 107.2 (1980): 135-143.

>>>>

\begin{abstract}
If $X$ is a compact Hausdorff space, an upper semi-continuous bonding function $f: X \rightarrow 2^{X}$ is said to be idempotent if $f^2 = f$. In this short paper, we prove that if $f: [0,1] \rightarrow C([0,1])$ is u.s.c., idempotent and surjective, but $f$ is not the identity, then the inverse limit with the single bonding function $f$ and with factor spaces indexed by a nonzero ordinal $\kappa$ contains a copy of $\kappa + 1$. It follows that such an inverse limit is only metric in the case where the index set $\kappa$ is countable.
\end{abstract}

@@

This short paper focuses on the case of an inverse limit with a single idempotent, surjective, upper semi-continuous, continuum-valued bonding function on $[0,1]$, when the index set of the inverse limit is some nonzero ordinal $\kappa$. In particular, we show that (so long as the bonding function is not the identity) such an inverse limit is metric exactly when $\kappa$ is countable. The strategy, as will be seen, is to show that the inverse limit must contain a copy of $\kappa + 1$.

@@

Let $\kappa$ be an ordinal. (We follow the convention of set theory in which an ordinal is equal to the set of its predecessors. For more detailed background on ordinals, see, for example, \cite{Kunen}. Assume all ordinals are nonzero.) Suppose that for each $\alpha \in \kappa$, $X_{\alpha}$ is a non-empty compact Hausdorff space and whenever $\alpha \le \beta \in \kappa$, the set-valued function $f_{\alpha \beta} : X_{\beta} \rightarrow 2^{X_{\alpha}}$ is u.s.c. (With this notation, $f_{\alpha \alpha}$ always denotes the identity on $X_{\alpha}$.) Suppose further that, whenever $\alpha \le \beta \le \eta \in \kappa$, we have $f_{\alpha \beta} \circ f_{\beta \eta} = f_{\alpha \eta}$. Then the collection $\textbf{f} = \{ f_{\alpha \beta} \ | \ \alpha \le \beta \in \kappa\}$ of u.s.c. functions is said to be \emph{exact}. In this case, we say $\{X_{\alpha}, f_{\alpha \beta}, \kappa\}$ is an \emph{inverse limit system}; the \emph{inverse limit}, $\varprojlim \{X_{\alpha}, f_{\alpha \beta}, \kappa\}$, of this system is $\{ \textbf{x} \in \prod_{\alpha \in \kappa} X_{\alpha} \ | \ x_{\alpha} \in f_{\alpha \beta}(x_{\beta})$ for all $\alpha \le \beta \in \kappa\}$. (In keeping with tradition, a boldface $\textbf{x}$ always denotes an element of $\prod_{\alpha \in \kappa} X_{\alpha}$; the $\alpha$th coordinate of such an $\textbf{x}$ will be denoted $\textbf{x}(\alpha)$.)  We call the u.s.c. functions $f_{\alpha \beta}$ the \emph{bonding functions} of the inverse limit, and each $X_{\alpha}$ is called a \emph{factor space} of the inverse limit.

@@

\textbf{Case 3:}
$x \in (0,1)$. Since $f(0) \subseteq (0,1]$ and the graph of $f$ is connected, $f(z) \subseteq (z,1]$ for each $z \in [0,x)$; thus, $0 \not\in f([0,x))$. Likewise,  since $f(1) \subseteq [0,1)$, we have $f(z) \subseteq [0,z)$ for each $z \in (x,1]$; thus, $1 \not\in f((x,1])$. Since $f$ is surjective,
$1\in f(v)$ for some $v\in[0,x]$ and $0\in f(w)$ for some $w\in[x,1]$. Let us say $f(v) = [k,1]$.

@@

So, $f(1) = [a,b]$ where $[a,b] \subseteq [0,x)$ or $[a,b] \subseteq (x,1)$. If $[a,b] \subseteq (x,1)$, then since $a > x$, $f(a) \subseteq [0,a)$ while $f(a)\subseteq f([a,b])=f(f(1))=f(1)=[a,b]$, contradiction. On the other hand, if $[a,b] \subseteq [0,x)$, then since $b<x$, $f(b) \subseteq (b,1]$ while $f(b) \subseteq f([a,b])=f(f(1))=f(1)=[a,b]$, another contradiction. Since each possible case results in a contradiction, the proof is complete.
\end{proof}

@@

\begin{proof}
As $G(f)$ satisfies $\Gamma$, let us say without loss of generality that $\<x,x\>, \<y,y\>$ and $\<x,y\>$ are elements of $G(f)$. $\varprojlim\{[0,1],f,\kappa\}$ contains the points $\textbf{x}_\alpha$ for $\alpha\leq\kappa$, defined by $\textbf{x}_\alpha(\beta)=y$ for $\beta<\alpha$ and $\textbf{x}_\alpha(\beta)=x$ otherwise. It is easy to see that $\{\textbf{x}_\alpha:\alpha\leq\kappa\}$ is a copy of $\kappa+1$.
\end{proof}

@@

A \textit{Corson compact} space is a compact space which embeds in a $\Sigma$-product of real lines (see \cite{alster}). All compact metric spaces are embeddable in $\mathbb R^\omega$ and therefore Corson compact.

@@

\bibitem{alster} Alster, K., and Pol, Roman, \textit{On function spaces of compact subspaces of $\Sigma$-products of the real line}, Fundamenta Mathematicae 107.2 (1980): 135-143.
